\name: A hierarchical Bayesian model for Supernova Cosmology

### 2018-10-10 papers/methods/paper.tex
Dan and Anais's feedback

In attempts to tease out the nature of dark energy, active and planned surveys are {\green continually growing in size and scale}. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has observed thousands of type Ia supernovae, attaining both spectroscopically and photometrically identified samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will observe scores of thousands of photometrically classified supernovae. {\green Such increased statistical power demands greater fidelity and flexibility in modelling the supernovae for cosmological purposes, as we will required reduced systematic uncertainties to fully utilised increase statistics} \citep{Betoule2014, Scolnic2017}.

As such, {\green considerable resources are} being put into developing more sophisticated supernova cosmology analyses. The role of simulations mimicking survey observations has become increasingly important in determining biases in cosmological constraints and validating specific supernova models. First used in SNLS \citep{Astier2006} and ESSENCE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, simulations are now a fundamental component of modern supernova cosmology.  \citet{Betoule2014} quantise and then correct observational bias using simulations, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulations to quantify observational bias in SN Ia distances as a function of {\green multiple} factors to better remove biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Bayesian Hierarchical models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, {\green and either use simulation-determined distance-corrections} or attempt to find sufficient analytic approximations for complicated effects such as Malmquist bias.

@@

Whilst supernova observations take the form of time-series photometric measurements of brightness in many photometric bands {\green and a redshift measurement of the supernova (or its assumed host)}, most analyses do not work from these measurements directly. Instead, most techniques fit an observed redshift and these photometric observations to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernova light curves for cosmology \citep{Guy2010, Betoule2014}. The resulting output from the model is, for each supernova, an amplitude $x_0$ (which can be converted into apparent magnitude, $m_B = -2.5\log(x_0)$), a stretch term $x_1$, and color term $c$, along with a covariance matrix describing the uncertainty on these summary statistics ($\mathcal{C}$). As all supernovae are not identical, an ensemble of supernovae form a redshift-dependent, observed population of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$, where the hat denotes an observed variable.

This {\green ensemble of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$} represents an observed population, which -- due to the presence of various selection effects -- may not represent the true, underlying supernova population. Accurately characterising this underlying population, its evolution over redshift, and effects from {\green host-galaxy} environment, is one of the challenges of supernova cosmology. Given some modelled underlying supernova population that lives in the redshift-dependent space $M_B$ {\green(absolute magnitude of the supernova, traditional in the Bessell $B$ band)}, $x_1$, and $c$, the introduction of cosmology into the model is simple -- it translates the underlying population in absolute magnitude space into the observed population in apparent magnitude space. Specifically, for any given supernova our map between absolute magnitude and apparent magnitude may take the traditional form:

@@

Sitting between the traditional models simplicity and the complexity of forward modelling lies Bayesian hierarchical models (BHM). Hierarchical models utilise multiple layers of connected parameters in their models, with the layers linked via well defined and {\green physically} motivated conditional probabilities. For example, an observation of a parameter from a population will be conditioned on the true value of the parameter, which itself will be conditioned on the population distribution of that parameter. We can thus easily incorporate different distributions, populations, and parameter inter-dependence which cannot be found in traditional analyses where uncertainty must be encapsulated in a covariance matrix.

@@

For the moment, let us consider a single supernova and its classic summary statistics $m_B$, $x_1$, $c$. For convenience, let us define $\eta \equiv \lbrace m_B, x_1, c \rbrace$. A full treatment of the summary statistics would involve determining $p(\hat{y}|\eta)$, where $\hat{y}$ represents the observed light curves {\green fluxes} and uncertainty. However, this is computationally prohibitive, and as such we rely on initially fitting the light curve observations to produce a best fit $\hat{\eta}$ along with a $3\times3$ covariance matrix $\mathcal{C}$ describing the uncertainty on $\hat{\eta}$. When using this simplification, our latent variables are given by
\begin{align}
p(\hat{\eta}|\eta) \sim \mathcal{N}(\hat{\eta} | \eta, \mathcal{C}). \label{eq:pop}
\end{align}
As discussed in Section \ref{sec:challenges},{\green the SALT2 model does not include full treatment of intrinsic dispersion in the supernova model, and thus this approximation does not encapsulate the full uncertainty introduced from this dispersion.}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{DES3YR_DES_BHMEFF_AMG10.pdf}
		\includegraphics[width=\columnwidth]{DES3YR_LOWZ_BHMEFF_G10.pdf}
	\end{center}
	\caption{Fitting the selection function for both the DES 3YR spectroscopically confirmed supernova sample and the combined low-redshift sample. Blue errorbars represent the efficiency calculated by determining the ratio of discovered to generated supernovae in apparent magnitude bins for SNANA simulations. The black line represents the best fit analytic function for each sample, and the light grey lines surrounding the best fit value represent random realisations of analytic function taking into account uncertainty on the best fit value.}
	\label{fig:sf_snana}
\end{figure}

\begin{table*}
	\centering
	\caption{Standardisation parameters and base intrinsic scatter parameter results for the 100 fits to {\gten} and {\celeven} simulations. We show the average parameter mean and {\green average }standard deviation respectively, with the simulation scatter shown in brackets. The width of the intrinsic scatter ($\sigma_{m_B}$) does not have an input truth value as it is determined from the scatter model, with {\gten} having 70\% achromatic scatter and {\celeven} only having 25\% achromatic scatter.}
	\label{tab:standardisation}
	\begin{tabular}{l|rrrrr}
		\hline
		Model & $\alpha - \alpha_{\rm True}$ & $\beta - \beta_{\rm True}$ & $\langle M_B \rangle - \langle M_B \rangle_{\rm True}$ & $\sigma_{m_B}^{\rm DES}$ & $\sigma_{m_B}^{{\rm low-}z}$ \\
		\hline
		{\gten} Stat + Syst  & $0.022, \ 0.009\ (0.008)$ & $ 0.34, \ 0.19\ (0.18)$ & $-0.002, \ 0.028\ (0.015)$ & $0.070, \ 0.022\ (0.018)$ & $0.073, \ 0.025\ (0.022)$\\
		{\gten}  Stat         & $0.000, \ 0.008\ (0.008)$ & $0.33, \ 0.20\ (0.17)$ & $0.001, \ 0.016\ (0.013)$ & $0.069, \ 0.023\ (0.019)$ & $0.072, \ 0.026\ (0.023)$\\
		{\celeven} Stat + Syst  &  $0.002, \ 0.009\ (0.007)$ & $-0.04, \ 0.15\ (0.13)$ & $-19.351, \ 0.030\ (0.018)$ & $0.024, \ 0.016\ (0.011)$ & $0.029, \ 0.020\ (0.014)$ \\
		{\celeven} Stat         &  $0.000, \ 0.008\ (0.007)$ & $-0.05, \ 0.16\ (0.13)$ & $-19.359, \ 0.016\ (0.015)$ & $0.025, \ 0.016\ (0.012)$ & $0.027, \ 0.020\ (0.015)$\\
		\hline
	\end{tabular}
\end{table*}

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. The quoted value for $w$ represents the average mean of the fits, with the uncertainty being the average standard deviation of the fits. In brackets, the simulation scatter (the standard deviation of the mean of 100 fits) is shown, and the bias significance represents our confidence that the deviation in the mean $w$ away from $-1$ is not due to statistical fluctuation. {\green The {\gten} model is consistent with $w=-1$, whilst the {\celeven} model show evidence of bias on $w$, scattering high.} However, their deviation from the truth value represents a shift of approximately $0.5\sigma$ when taking into account the uncertainty on fits to $w$. The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. We also note that the simulations do not vary cosmological parameters nor population. As our model does include uncertainty on those values, the simulation scatter is expected to be less than the model uncertainty, and represents a minimum bound on permissible uncertainty values.}
	\label{tab:bulk_summary}
	\begin{tabular}{l|rr}
		\hline
		Model & $w\ \langle\mu\rangle,\ \langle\sigma_w\rangle$ (scatter) & $w$-Bias \\
		\hline
		{\gten} Stat + Syst  &  $-0.998, \ 0.097\ (0.073)$ & $(0.02\pm0.07)\sigma$  \\
		{\gten}  Stat         &  $-1.008,\ 0.080\ (0.068)$ & $(-0.10\pm0.08)\sigma$   \\
		{\celeven} Stat + Syst  &  $-0.945,\ 0.098\ (0.077)$ & $(0.55\pm0.08)\sigma$  \\
		{\celeven} Stat         &  $-0.948,\ 0.079\ (0.066)$ & $(0.65\pm0.08)\sigma$   \\
		\hline
	\end{tabular}
\end{table}

@@

\begin{table}
	\centering
	\caption{Characterising the bias on $w$ using the 100 simulations for the {\gten} scatter model and 100 simulations for {\celeven} scatter model. We also show the results when combining the {\gten} and {\celeven} models into a combined set of 200 simulations. The mean $w$ value for our method and BBC are presented, along with the mean when averaging the difference between our method and BBC for each individual simulation. Averages are computed giving each simulation sample the same weight. In the model, $\Delta$ represents Steve - BBC. The final row shows the scatter between Steve and BBC for the different simulations.}
	\label{tab:bulk_summary_bbc}
	\begin{tabular}{l|ccc}
		\hline
		Model & {\gten} & {\celeven} & ({\gten} + {\celeven}) \\
		\hline
		Steve $\langle w \rangle$           & $-0.998\pm0.007$ & $-0.945\pm0.007$ & $-0.972\pm0.006$\\
		BBC $\langle w \rangle$             & $-1.044\pm0.006$ & $-0.978\pm0.007$ & $-1.010\pm0.005$\footnotemark\\
		$\Delta$  $\langle w \rangle$  & $+0.044\pm0.006$ & $+0.033\pm0.006$ & $+0.038\pm0.004$ \\
		\hline
		$\Delta$  $\sigma_w$           & $0.057\pm0.004$ & $0.062\pm0.004$ & $0.060\pm0.003$ \\
		\hline
	\end{tabular}
\end{table}
\footnotetext{This value is computed with each simulation having the same weight. It disagrees with the value provided in \citet[Table 10, row 3]{Brout18SYS} which uses inverse variance weighted averages. We do not utilise this weight because the variance is correlated with the value of $w$ due to the $\Omega_m$ prior applied in the fitting process.} 
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{bulk_comparison.pdf}
	\end{center}
	\caption{Recovered $w$ for the 200 validation simulations with full treatment of statistical and systematic errors. Uncertainty on the recovered $w$ value is shown for every second data point for visual clarity.}
	\label{fig:bulk_comparison}
\end{figure}

>>>>

In attempts to tease out the nature of dark energy, active and planned surveys are {\green continually growing in size and scale}. The Dark Energy Survey \citep[DES,][]{Bernstein2012, Abbott2016} has {\green discovered} thousands of type Ia supernovae, attaining both spectroscopically and photometrically identified samples. The Large Synoptic Survey Telescope \citep[LSST,][]{Ivezic2008, LSSTScienceCollaboration2009} will {\green discover} scores of thousands of photometrically classified supernovae. {\green Such increased statistical power demands greater fidelity and flexibility in modelling supernovae for cosmological purposes, as we will require reduced systematic uncertainties to fully utilise these increased statistics} \citep{Betoule2014, Scolnic2017}.

As such, {\green considerable resources are} being put into developing more sophisticated supernova cosmology analyses. The role of simulations mimicking survey observations has become increasingly important in determining biases in cosmological constraints and validating specific supernova models. First used in SNLS \citep{Astier2006} and ESSENCE analyses \citep{Wood-Vasey2007}, and then refined and improved in \citet{Kessler2009}, simulations are now a fundamental component of modern supernova cosmology.  \citet{Betoule2014} quantise and then correct observational bias using simulations, and more recently \citet{Scolnic2016} and \citet{Kessler2017} explore simulations to quantify observational bias in SN Ia distances as a function of {\green multiple} factors to better remove biases. Approximate Bayesian computation methods also make use of simulations, trading traditional likelihoods and analytic approximations for more robust models with the cost of increased computational time \citep{Weyant2013, Jennings2016}. Bayesian Hierarchical models abound \citep{Mandel2009, March2011, March2014, Rubin2015, Shariff2016, Roberts2017}, {\green and either use simulation-determined distance-corrections} or attempt to find analytic approximations for complicated effects such as Malmquist bias.

@@

Whilst supernova observations take the form of photometric time-series brightness measurements in many bands {\green and a redshift measurement of the supernova (or its assumed host)}, most analyses do not work from these measurements directly. Instead, most techniques fit an observed redshift and these photometric observations to a supernova model, with the most widely used being that of the empirical SALT2 model \citep{Guy2007, Guy2010}. This model is trained separately before fitting the supernova light curves for cosmology \citep{Guy2010, Betoule2014}. The resulting output from the model is, for each supernova, an amplitude $x_0$ (which can be converted into apparent magnitude, $m_B = -2.5\log(x_0)$), a stretch term $x_1$, and color term $c$, along with a covariance matrix describing the uncertainty on these summary statistics ($\mathcal{C}$). As all supernovae are not identical, an ensemble of supernovae form a redshift-dependent, observed population of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$, where the hat denotes an observed variable.

This {\green ensemble of $\hat{m}_B$, $\hat{x}_1$ and $\hat{c}$} represents an observed population, which -- due to the presence of various selection effects -- may not represent the true, underlying supernova population. Accurately characterising this underlying population, its evolution over redshift, and effects from {\green host-galaxy} environment, is one of the challenges of supernova cosmology. Given some modelled underlying supernova population that lives in the redshift-dependent space $M_B$ {\green(absolute magnitude of the supernova, traditionally in the Bessell $B$ band)}, $x_1$, and $c$, the introduction of cosmology into the model is simple -- it translates the underlying population from absolute magnitude space into the observed population in apparent magnitude space. Specifically, for any given supernova our map between absolute magnitude and apparent magnitude may take the traditional form:

@@

Sitting between the traditional models simplicity and the complexity of forward modelling lies Bayesian hierarchical models (BHM). Hierarchical models utilise multiple layers of connected parameters, with the layers linked via well defined and {\green physically} motivated conditional probabilities. For example, an observation of a parameter from a population will be conditioned on the true value of the parameter, which itself will be conditioned on the population distribution of that parameter. We can thus easily incorporate different distributions, populations, and parameter inter-dependence which cannot be found in traditional analyses where uncertainty must be encapsulated in a covariance matrix.

@@

For the moment, let us consider a single supernova and its classic summary statistics $m_B$, $x_1$, $c$. For convenience, let us define $\eta \equiv \lbrace m_B, x_1, c \rbrace$. A full treatment of the summary statistics would involve determining $p(\hat{y}|\eta)$, where $\hat{y}$ represents the observed light curves {\green fluxes} and uncertainty. However, this is computationally prohibitive, and as such we rely on initially fitting the light curve observations to produce a best fit $\hat{\eta}$ along with a $3\times3$ covariance matrix $\mathcal{C}$ describing the uncertainty on $\hat{\eta}$. When using this simplification, our latent variables are given by
\begin{align}
p(\hat{\eta}|\eta) \sim \mathcal{N}(\hat{\eta} | \eta, \mathcal{C}). \label{eq:pop}
\end{align}
As discussed in Section \ref{sec:challenges}, {\green the SALT2 model does not include full treatment of intrinsic dispersion in the supernova model, and thus this approximation does not encapsulate the full uncertainty introduced from this dispersion.}

@@

\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{DES3YR_DES_BHMEFF_AMG10.pdf}
		\includegraphics[width=\columnwidth]{DES3YR_LOWZ_BHMEFF_G10.pdf}
	\end{center}
	\caption{Fitting the selection function for both the DES 3YR spectroscopically confirmed supernova sample and the low-$z$ sample. Blue errorbars represent the efficiency calculated by determining the ratio of discovered to generated supernovae in apparent magnitude bins for SNANA simulations. The black line represents the best fit analytic function for each sample, and the light grey lines surrounding the best fit value represent random realisations of analytic function taking into account uncertainty on the best fit value.}
	\label{fig:sf_snana}
\end{figure}

\begin{table*}
	\centering
	\caption{Standardisation parameters and base intrinsic scatter parameter results for the 100 fits to {\gten} and {\celeven} simulations. We show the average parameter mean and {\green average }standard deviation respectively, with the simulation scatter shown in brackets, such that each cell shows $\langle\mu\rangle\ [\langle\sigma\rangle\ ({\rm scatter})]$. The width of the intrinsic scatter ($\sigma_{m_B}$) does not have an input truth value as it is determined from the scatter model, with {\gten} having 70\% achromatic scatter and {\celeven} only having 25\% achromatic scatter. A measure of the significance of the parameter bias can be calculated by comparing the bias to a tenth of the scatter (as our Monte-Carlo estimate uncertainty is $\sqrt{100}$ of the scatter). From this, we can see that most biases are detected with high statistical significance due to the large number of simulations tested against.}
	\label{tab:standardisation}
	\begin{tabular}{l|rrrrr}
		\hline
		Model & $\alpha - \alpha_{\rm True}$ & $\beta - \beta_{\rm True}$ & $\langle M_B \rangle - \langle M_B \rangle_{\rm True}$ & $\sigma_{m_B}^{\rm DES}$ & $\sigma_{m_B}^{{\rm low-}z}$ \\
		\hline
		{\gten} Stat + Syst  & $0.022 \ [0.009\ (0.008)]$ & $ 0.34 \ [0.19\ (0.18)]$ & $-0.002 \ [0.028\ (0.015)]$ & $0.070 \ [0.022\ (0.018)]$ & $0.073 \ [0.025\ (0.022)]$\\
		{\gten}  Stat         & $0.000 \ [0.008\ (0.008)]$ & $0.33 \ [0.20\ (0.17)]$ & $0.001 \ [0.016\ (0.013)]$ & $0.069 \ [0.023\ (0.019)]$ & $0.072 \ [0.026\ (0.023)]$\\
		{\celeven} Stat + Syst  &  $0.002 \ [0.009\ (0.007)]$ & $-0.04 \ [0.15\ (0.13)]$ & $0.014 \ [0.030\ (0.018)]$ & $0.024 \ [0.016\ (0.011)]$ & $0.029 \ [0.020\ (0.014)]$ \\
		{\celeven} Stat         &  $0.000 \ [0.008\ (0.007)]$ & $-0.05 \ [0.16\ (0.13)]$ & $0.006 \ [0.016\ (0.015)]$ & $0.025 \ [0.016\ (0.012)]$ & $0.027 \ [0.020\ (0.015)]$\\
		\hline
	\end{tabular}
\end{table*}

@@

\begin{table}
	\centering
	\caption{Investigating the combined 100 fits to {\gten} and {\celeven} simulations, fitting with both statistics only and also when including systematics. The quoted value for $w$ represents the average mean of the fits, with the average uncertainty being shown in square brackets and the simulation scatter (the standard deviation of the mean of 100 fits) shown in standard brackets. The bias significance represents our confidence that the deviation in the mean $w$ away from $-1$ is not due to statistical fluctuation. {\green The {\gten} model is consistent with $w=-1$, whilst the {\celeven} model show evidence of bias on $w$, scattering high.} However, their deviation from the truth value represents a shift of approximately $0.5\sigma$ when taking into account the uncertainty on fits to $w$. The bias is sub-dominant to both the size of the uncertainty for each fit, and the scatter induced by statistical variance in the simulations. We also note that the simulations do not vary cosmological parameters nor population. As our model does include uncertainty on those values, the simulation scatter is expected to be less than the model uncertainty, and represents a minimum bound on permissible uncertainty values.}
	\label{tab:bulk_summary}
	\begin{tabular}{l|rr}
		\hline
		Model & $w\ \langle\mu\rangle\ [\langle\sigma_w\rangle$ (scatter)$]$ & $w$-Bias \\
		\hline
		{\gten} Stat + Syst  &  $-0.998 \ [0.097\ (0.073)]$ & $(0.02\pm0.07)\sigma$  \\
		{\gten}  Stat         &  $-1.008\ [0.080\ (0.068)]$ & $(-0.10\pm0.08)\sigma$   \\
		{\celeven} Stat + Syst  &  $-0.945\ [0.098\ (0.077)]$ & $(0.55\pm0.08)\sigma$  \\
		{\celeven} Stat         &  $-0.948\ [0.079\ (0.066)]$ & $(0.65\pm0.08)\sigma$   \\
		\hline
	\end{tabular}
\end{table}

@@

\begin{table}
	\centering
	\caption{Characterising the bias on $w$ using the 100 simulations for the {\gten} scatter model and 100 simulations for {\celeven} scatter model. We also show the results when combining the {\gten} and {\celeven} models into a combined set of 200 simulations. The mean $w$ value for our method and BBC are presented, along with the mean when averaging the difference between our method and BBC for each individual simulation. Averages are computed giving each simulation sample the same weight. In the model, $\Delta$ represents Steve - BBC. The final row shows the scatter between Steve and BBC for the different simulations.}
	\label{tab:bulk_summary_bbc}
	\begin{tabular}{l|ccc}
		\hline
		Model & {\gten} & {\celeven} & ({\gten} + {\celeven}) \\
		\hline
		Steve $\langle w \rangle$           & $-0.998\pm0.007$ & $-0.945\pm0.007$ & $-0.972\pm0.006$\\
		BBC $\langle w \rangle$             & $-1.044\pm0.006$ & $-0.978\pm0.007$ & $-1.010\pm0.005$\footnotemark\\
		$\Delta$  $\langle w \rangle$  & $+0.044\pm0.006$ & $+0.033\pm0.006$ & $+0.038\pm0.004$ \\
		\hline
		$\Delta$  $\sigma_w$           & $0.057\pm0.004$ & $0.062\pm0.004$ & $0.060\pm0.003$ \\
		\hline
	\end{tabular}
\end{table}
\footnotetext{This value is computed with each simulation having the same weight. It disagrees with the value provided in \citet[Table 10, row 3]{Brout18SYS} which uses inverse variance weighted averages. We do not utilise this weight because the variance is correlated with the value of $w$ due to the $\Omega_m$ prior applied in the fitting process. We note that if inverse variance weighting is applied to both datasets, they both shift by $\Delta w \approx 0.01$, and thus the predicted difference between the BBC method and {\steve} remains the same.} 
\begin{figure}
	\begin{center}
		\includegraphics[width=\columnwidth]{bulk_comparison.pdf}
	\end{center}
	\caption{Recovered $w$ for the 200 validation simulations with full treatment of statistical and systematic errors. Uncertainty on the recovered $w$ value is shown for every second data point for visual clarity.}
	\label{fig:bulk_comparison}
\end{figure}
